---
title: "Application of phyloAMR ancestral state reconstruction to evaluate phylogenetics of ermC"
author: "Kyle Gontjes"
date: "`r Sys.Date()`"
output: html_document
---

# Purpose: Evaluate phylogenetics of ermC plasmid in MRSA dataset
# 
# Methodology: Ancestral state reconstruction of ermC plasmid trait using phyloAMR (https://github.com/kylegontjes/phyloAMR)
## 1. Identified ancestral states using phyloAMR's asr() function 
## 2. Identified phylogenetic events of ermC plasmids using phyloAMR's asr_cluster_detection() function
## 3. Classified phylogenetic events of ermC plasmids using phyloAMR's asr_cluster_analysis() function
## 4. Visualization of clustering across the phylogeny using ggtree 
## 5. Visualization of clustering across the study period using ggplot2  
# 
# Working directory: /nfs/turbo/umms-esnitkin-secure/Project_MRSA/Analysis/2020-MRSA_jail_col_vs_inf/2025-05-29_phyloAMR
## Tree: /nfs/turbo/umms-esnitkin-secure/Project_MRSA/Analysis/2020-MRSA_jail_col_vs_inf/2025-05-29_phyloAMR/combined_tree.treefile
## Minimal datset: /nfs/turbo/umms-esnitkin-secure/Project_MRSA/Analysis/2020-MRSA_jail_col_vs_inf/2025-05-29_phyloAMR/combined_tree_metadata.csv
## Code, final results, and figure in /nfs/turbo/umms-esnitkin-secure/Project_MRSA/Analysis/2020-MRSA_jail_col_vs_inf/2025-05-29_phyloAMR/thiede_steinberg_et_al_MRSA_ermC/
# 
# Github repo: https://github.com/sthiede/thiede_steinberg_et_al_MRSA_ermC/

```{r setup, include=FALSE}
require(knitr)
knitr::opts_chunk$set(echo = TRUE)
opts_knit$set(echo = T,warning = FALSE, message = FALSE,error=F,comment=NA) 
opts_chunk$set(echo = T,warning = FALSE, message = FALSE,error=F,comment=NA)
opts_knit$set(root.dir='~/Desktop/armis_mount/')
```

# Environment
## The analyses was performed on Kyle Gontjes' core imaged Mac, issued by University of Michigan 
```{r}
#Packages
packages <- c("tidyverse",'ape','ggtree','phyloAMR','lubridate','kableExtra','ggtext')   
#Load Packages
invisible(lapply(packages,library,character.only=T,quietly=T))  
# Seed
set.seed(45)
# Print
Sys.info()
sessionInfo()  
```

## Visualization function for tables
```{r}
favorite_kable <- function(x){
    x %>% kable(., format = "html", table.attr = "style='width:100%;'", row.names = F) %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
}
```

# Load data
## 1. Data stored in /nfs/turbo/umms-esnitkin-secure/Project_MRSA/Analysis/2020-MRSA_jail_col_vs_inf/2025-05-29_phyloAMR
### a. combined_tree.treefile: Phylogeny
### b. combined_tree_metadata.csv: Dataframe with ermc, seq_id, patient_id, and collect_date variables
```{r}
tr <- read.tree("./combined_tree.treefile")
df <- read.csv("./combined_tree_metadata.csv")
df <- as.data.frame(df)
```

# Curate the data 
```{r}
# Converted collect_date into a date
df$date <- lubridate::as_date(df$collect_date)

# Bin the dates by month to enable easy visualization in epicurve
df$date_bin <- floor_date(df$date,'month') 

# Add rownames to the dataframe
rownames(df) <- df$seq_id
```

# Curate the phylogeny
```{r}
# Midpoint phylogeny
tr_midpoint <- phytools::midpoint.root(tr)
tr_obj <- ggtree(tr_midpoint)

# Match tree ordering
df <-  df[match(tr_midpoint$tip.label, df[['seq_id']]), ]
```

# Get ancestral state reconstruction of ermC plasmid
## Generate using corHHM's joint ancestral state reconstruction algorithm
```{r}
ermC_asr <- phyloAMR::asr(df = df, tr = tr_midpoint, tip_name_variable = "seq_id",trait = 'ermc', model = "MF", node_states = 'joint')

saveRDS(ermC_asr,"./thiede_steinberg_et_al_MRSA_ermC/data/phylogenetic_analysis/ermC_ASR_object.RDS")
```

## Model statistics
### phyloAMR's model finder optioned was performed to utilize the rate matrix with the best fit
```{r}
ermC_asr$corHMM_model_statistics %>% favorite_kable()

write_csv(ermC_asr$corHMM_model_statistics,"./thiede_steinberg_et_al_MRSA_ermC/data/phylogenetic_analysis/ermC_ASR_model_statistics.csv")
```

## Transition statistics
### The number of transitions and continuation events across the edges of the phylogeny

```{r}
transition_stats <- phyloAMR::asr_transition_analysis(ermC_asr$parent_child_df, node_states = "joint")

transition_stats %>% favorite_kable()

write_csv(transition_stats,"./thiede_steinberg_et_al_MRSA_ermC/data/phylogenetic_analysis/ermC_ASR_transition_statistics.csv")
``` 

# Phylogenetic clustering of ermC plasmid using ancestral state reconstruction data
## Used phyloAMR's ancestral state reconstruction data and asr_cluster_detection() algorithm to identify isolates that shared ancestral gain events
```{r}
ermC_asr_cluster <- asr_cluster_detection(df = df, tr = tr_midpoint, tip_name_variable = "seq_id",patient_id = 'patient_id', parent_child_df = ermC_asr$parent_child_df, node_states = 'joint', simplify_faux_clusters = TRUE, simplify_revertant = TRUE, collapse_cluster =  TRUE, confidence = NULL)

write_csv(ermC_asr_cluster,"./thiede_steinberg_et_al_MRSA_ermC/data/phylogenetic_analysis/ermC_ASR_clustering.csv")
```

## Clustering statistics for ermC
### The number of phylogenetic events (i.e., singletons and clusters)

```{r}
cluster_stats <- asr_cluster_analysis(ermC_asr_cluster)

cluster_stats %>% favorite_kable()

write_csv(cluster_stats,"./thiede_steinberg_et_al_MRSA_ermC/data/phylogenetic_analysis/ermC_ASR_clustering_statistics.csv")
```

# Figure A - Phylogeny with clustering
## Clustering scale
```{r}
ncluster <- cluster_stats$clusters

color_palette <- grDevices::colors() %>% subset(grepl("red",.)==F)
clusters_col <- hues::iwanthue(ncluster,hmin=13,hmax=360,lmin = 0,lmax = 100,cmin=0,cmax=180,random = F,plot = )
clusters_name <- paste0("Cluster ",1:ncluster)
names(clusters_col) <- clusters_name

cluster_scale <- scale_fill_manual(breaks =c("No feature","Singleton",names(clusters_col)),values=c("No feature" = "#A9A9A9","Singleton" = "red",clusters_col),labels=c("Absent","Singleton",clusters_name),name="Phylogenetics of the *ermC* plasmid", guide = guide_legend(order=2,ncol=8, title.position = "top", label.position = "right"))
```
 
## Painted tree
### We painted the edges with the presence (red) or absence (black) of the plasmid
```{r}
custom_theme <- scale_color_manual(breaks = c(1,0), labels = c("Present","Absent"), values= c("red","black"), name = '*ErmC* Plasmid',guide = guide_legend(title.position = "top", label.position = "right",nrow=2,order=1))

get_painted_tree_with_states <- function(parent_child_df,tr){
    num_nodes <- nrow(tr$edge) + 1
    root_node <- length(tr$tip.label) + 1
    root_value <- unique(parent_child_df[parent_child_df$parent == 
        root_node, "parent_value"])
    parent_child_df <- parent_child_df[order(parent_child_df$child, 
        decreasing = FALSE), ]
    number_tips <- length(tr$tip.label)
    outcome_str <- parent_child_df[parent_child_df$child <= number_tips, 
        "child_value"]
    parent_str <- parent_child_df[parent_child_df$child > number_tips, 
        "child_value"]
    state_df <- data.frame(node = seq_len(num_nodes), state = c(outcome_str, 
        root_value, parent_str))
    tr <- full_join(tr, state_df)
    painted_tree <- ggtree(tr, mapping = aes(color = as.factor(state)),size=0.3) + 
        custom_theme
    return(painted_tree)
}

p1 <- get_painted_tree_with_states(ermC_asr$parent_child_df,tr = tr_midpoint) + custom_theme + theme(legend.position = 'bottom',legend.title = ggtext::element_markdown())

p1
```

## Painted phylogeny with clustering data
```{r,fig.height=14,fig.width=15}
# Theme
consistent_theme <- theme(legend.position = 'bottom',legend.direction="horizontal", legend.justification = "center", legend.key = element_rect(colour = c('black')),legend.box.spacing = unit(.0001, "cm"),legend.key.size = unit(.75, "cm"),legend.key.width = unit(.75, "cm"),legend.spacing.x=unit(.75, "cm"), legend.title = ggtext::element_markdown(size=20,hjust=0.5),legend.text = element_text(size=18,hjust=0),legend.spacing = unit(0,'cm'),legend.margin = margin(t=-1,b=0.25,unit='cm'))

# Add new scale fill
p1.1 <- p1 + ggnewscale::new_scale_fill()

# Add matrix with cluster calls
A <- gheatmap(p1.1, ermC_asr_cluster %>% select(asr_cluster_renamed)  %>% `colnames<-`("Plasmid clustering") %>% mutate_all(as.factor) , colnames_position = "top",colnames_angle=90, colnames_offset_y = 0.25, hjust = 0, color = NA, font.size = 6, width = .1,offset =0) + cluster_scale + ylim(NA,1250) + theme(legend.position = 'bottom') + consistent_theme
 
A
```

# Figure B - Histogram with clustering over time
## Themes
```{r, fig.height=7,fig.width=7.5}
format <- theme(legend.position = "bottom",
                  axis.text =   element_text(size=18, color="black"),
                  axis.title = element_text(size = 20, color="black"),
                  axis.ticks.length = unit(0.25, "cm"),
                  legend.text =   element_text(size=18, color="black"),
                  legend.title = element_text(size = 20, color="black")
)

theme_bw_me <- theme(panel.background = element_rect(fill = "white",colour = NA), panel.grid = element_blank(),
                     strip.background = element_rect(fill = "white",colour = "black"),
                     panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     axis.line = element_line(colour = "black"),legend.position = "bottom")
```

## Add clustering data to metadata
```{r, fig.height=7,fig.width=7.5}
df <- left_join(df,ermC_asr_cluster %>% mutate(seq_id = tip_name)) 

cluster_levels <- c(paste0("Cluster ",ncluster:1),"Singleton","No feature")

df$asr_cluster_renamed_rev <- factor(df$asr_cluster_renamed, levels = cluster_levels,labels = gsub("No feature","Absent",cluster_levels))
```

## Histogram
```{r, fig.height=7,fig.width=7.5}
x_year_scale <-scale_x_date(date_breaks = "1 year",date_labels = "%Y")

B <- ggplot(data = df, aes(x = date_bin, fill = fct_rev(fct_infreq(asr_cluster_renamed_rev)))) + geom_histogram(stat  = 'count') + format + theme_bw_me + theme(legend.position='none')  + xlab("Time (months)") + ylab("Number of isolates") +  cluster_scale + x_year_scale

B
```

# Figure C - Duration of clusters in sample collection
## Cluster isolate, patient, statistics  statistics
```{r}
groups <- c("Absent","Singleton",names(clusters_col))

get_cluster_stats <- function(group){
  df_sub <- df %>% subset(asr_cluster_renamed_rev == group)
  n_isolates <- as.numeric(nrow(df_sub))
  n_patients <- as.numeric(length(unique(df_sub$patient_id)))
  subsequent_transmission <- as.numeric(n_patients - 1)
  
  sum <- as.data.frame(summary(df_sub$date)) %>% t %>% as.data.frame %>% `colnames<-`(snakecase::to_snake_case(colnames(.))) %>% mutate_all(lubridate::as_date)
  sum$duration <- sum$max - sum$min 
  cbind(group,n_isolates,n_patients,subsequent_transmission,sum) %>% `rownames<-`(NULL)
}

cluster_individual_data <- lapply(groups,get_cluster_stats) %>% do.call(rbind.data.frame,.) 

cluster_individual_data %>% favorite_kable()

write_csv(cluster_individual_data,"./thiede_steinberg_et_al_MRSA_ermC/data/phylogenetic_analysis/ermC_ASR_cluster_individual_data.csv")
```

## Differences in subsequent transmission events across clusters (i.e., n_patients - 1)
```{r}
cluster_only_individual_stats <- subset(cluster_individual_data,!group %in% c("Absent","Singleton"))

summary(cluster_only_individual_stats$subsequent_transmission)
```

## Cluster duration differences
```{r}
summary(cluster_only_individual_stats$duration)
```

## Cluster duration boxplot
```{r, fig.height=7,fig.width=7.5}
# Edit y-axis
groups <- levels(df$asr_cluster_renamed_rev)
cluster_number_name <- sapply(groups,FUN=function(group_name,cluster_individual_data){
  stats_sub <- subset(cluster_individual_data,group == group_name) 
  n_patients <- stats_sub[["n_patients"]]
  paste0(group_name," (n = ",n_patients,")") 
},cluster_individual_data = cluster_individual_data) %>% `names<-`(groups) 

# Recode y axis labels using cluster_number_name 
recoded_y_scale <-  scale_y_discrete(labels = cluster_number_name)

# Cluster duration in study
C <- ggplot(data = df, aes(x = date_bin, y = asr_cluster_renamed_rev, fill = asr_cluster_renamed_rev))  + format + geom_boxplot() + theme_bw_me  + theme(legend.position='none') + xlab("Time (months)") + ylab("")+  cluster_scale + recoded_y_scale

C
```

# Supplemental figure: Frequent emergence and spread of ermC plasmid-containing strains

```{r,fig.width=15,fig.height=20}
cow_A <- cowplot::plot_grid(A,labels="A",label_size=24)
cow_BC <- cowplot::plot_grid(B,C,ncol=2,labels=c("B","C"),label_size = 24)

final_figure <- cowplot::plot_grid(cow_A,cow_BC,ncol=1,labels = c("",""),rel_heights = c(1,0.50))

final_figure
```

```{r}
ggsave("./thiede_steinberg_et_al_MRSA_ermC/figures/s_figure_ermC_asr_clustering.png",plot = final_figure, dpi = 900, width = 15, height = 20,bg='white')
``` 